\LARGE\textbf{Designing Studies and Evaluating Research Results:\\ Type M and Type S Errors for Pearson Correlation Coefficient }

### 2020-05-05 Documents/Paper_main/Paper_main.Rnw
Renew bibliography

%----    Basic packages    ----%
\documentclass{article}
\usepackage[T1] {fontenc} 		  % Font encoding
\usepackage [utf8] {inputenc}		% Encoding for the document
\usepackage[a4paper,includeheadfoot,left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}  % margin settings
\usepackage[english]{babel}

@@

%----    Other packages    ----%
\usepackage{tikz}
\usepackage{graphicx}   % for including graphics
\usepackage{amsmath}    % for math equation
\usepackage{bm}         % for bold math
\usepackage{csquotes}
\usepackage{enumitem}   % for enumarate and itemize
\setlist[enumerate]{font=\bfseries} % bold font for enumerate list
\setlist[enumerate]{itemsep=0ex}    % reduce space between items
\setlist[itemize]{itemsep=0ex}      % reduce space between items
\usepackage[labelfont=bf]{caption}  % caption
\usepackage{lineno}                 % rows number
\usepackage[doublespacing]{setspace}% line spacing
\usepackage{titlesec}               % redefine title format for appendix
% \usepackage[style=apa, backend=biber]{biblatex}  % bibliografia
% \addbibresource{Paper_main.bib}
\usepackage[natbibapa]{apacite}
\bibliographystyle{apacite}

\usepackage[hyphens,spaces]{url}
\usepackage{hyperref}    % ref between elements

@@

Psychological science is increasingly committed to scrutinize its published findings by promoting large-scale replication efforts \citep{camererEvaluatingReplicabilityLaboratory2016, camererEvaluatingReplicabilitySocial2018, ebersoleManyLabsEvaluating2016, kleinInvestigatingVariationReplicability2014, kleinManyLabsInvestigating2018, opensciencecollaborationEstimatingReproducibilityPsychological2015}. In replication studies, the protocol of a previously published study is repeated as closely as possible with a new sample. Interestingly, many replication studies found smaller effects than originals, suggesting a sort of decline effect from original studies to replications \citep{camererEvaluatingReplicabilitySocial2018, opensciencecollaborationEstimatingReproducibilityPsychological2015, schoolerTurningLensScience2014}. Among the many factors that could explain such decline, one relates to a feature of study design: statistical power \citep{protzkoDeclineEffectsTypes2017}. In particular, it is plausible for original studies to have lower statistical power than their replications. When underpowered studies are analyzed using thresholds, such as statistical significance levels, effects passing such thresholds have to exaggerate the true effect size \citep{buttonPowerFailureWhy2013, gelmanTypeErrorMight2017, ioannidisWhyMostDiscovered2008, ioannidisEmergenceLargeTreatment2013, laneEstimatingEffectSize1978}. Indeed, as will be extensively shown below, in underpowered studies only large effects correspond to values that can reject the null hypothesis and be statistically significant. As a consequence, if the original study was underpowerd and found an exaggerated estimate of the effect, the replication results will likely show a decline. The concept of statistical power finds its natural developement in the Neyman-Pearson framework of statistical inference and this is the framework that we adopt in this contribution. Further discussion on the Neyman and Perason approach and a comparison with Null Hypothesis Significance Testing (NHST) is available in \citet{altoeEnhancingStatisticalInference2020} and \citet{gigerenzerNullRitualWhat2004}.

Effect size exaggeration is a corollary consequence of low statistical power that might not be evident at first. Indeed, by definition, statistical power focuses on statistical significance, not effect size estimation (i.e., statistical power is the probability to find a statistically significant result if an effect of a certain size actually exists). However, effect size estimation and statistical significance are closely related. This point can be highlighted estimating the Type M (magnitude) and Type S (sign) errors characterizing a study design \citep{gelmanPowerCalculationsAssessing2014}. Given a study design (i.e., sample size, statistical test directionality, and $\alpha$ level), Type M error indicates the factor by which a statistically significant effect would be, on average, exaggerated. Type S error indicates the probability to find a statistically significant effect in the opposite direction to the one considered plausible. It is evident that both errors are defined starting from a reasoned guess on the plausible magnitude and direction of the effect under study, which is called \emph{plausible effect size} \citep{gelmanPowerCalculationsAssessing2014}. The analysis that researchers perform to consider Type M and Type S errors in their research practice is called \emph{design analysis}, given the special focus posed into considering the design of a study \citep{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

Why do these errors matter? The exaggeration of effect sizes, in the right or in the wrong direction, has important consequences on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, which could turn out to be underpowered if they do not take into account possible inflations of the original effect \citep{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \citep{gelmanEmbracingVariationAccepting2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

Type M (magnitude) and Type S (sign) errors are not widely known in the psychological research community but their consideration during the research process has the potential to improve current research practices, for example, by increasing the awarness that design choices have on possible studies' results. In a previous work we illustrated Type M and Type S errors using Cohens’\emph{d} as a measure of effect size \citep{altoeEnhancingStatisticalInference2020}. In this contribution we aim to to further increase familiarity with Type M and Type S errors, considering another common effect size measures in psychology: Pearson correlation coefficient, $\rho$.  In the following paragraphs we discuss what Type M and Type S errors are and how to perform a design analysis using the ad-hoc R functions that we developed.

@@

Pearson correlation coefficient is a standardized effect size measure indicating the strength and the direction of the relationship between two continuous variables \citep{cohenStatisticalPowerAnalysis1988, ellisEssentialGuideEffect2010}. Even though the correlation coefficient is widely known, we briefly go over its main features using an example. Imagine that we were interested to measure the relationship between anxiety and depression in a population and we plan a study with \emph{n} participants, where, for each participant, we measure the level of anxiety (i.e., variable X) and the level of depression (i.e., variable Y). At the end of the study we will have n pairs of values X and Y. The correlation coefficient helps us answer the questions: how strong is the relationship between anxiety and depression in this population? Is the relationship positive or negative? Correlation ranges from -1 to +1, indicating respectively two extreme scenarios of perfect negative relationship and perfect positive relationship. Since the correlation coefficient is a dimensionless number, it is a signal to noise ratio where the signal is given by the covariance between the two variables ($cov(x,y)$) and the noise is expressed by the product between the standard deviations of the two variables ($S_x  S_y$; see Formula\ref{eq:correlation}). Following the conventional standards, in this contribution we will use the symbol $\rho$ to indicate the correlation in the population and the symbol \emph{r} to indicate the value measured in a sample.
\begin{equation}
\label{eq:correlation}
r=\frac{cov(x,y)}{S_x S_y}.
\end{equation}

Magnitude and sign are two important features characterizing Pearson correlation coefficient and effect size measures in general. And, when estimating effect sizes, errors could be committed exactly regarding these two aspects. \citet{gelmanPowerCalculationsAssessing2014} introduced two indexes to quantify these risks:
\begin{itemize}
  \item{Type M error, where M stands for magnitude, is also called Exaggeration Ratio - the factor by which a statistically significant effect is on average exaggerated.}
  \item{Type S error, sign - the probability to find a statistically significant result in the opposite direction to the plausible one.}
\end{itemize}

How are these errors computed? In the next paragraphs we approach this question preferring an intuitive perspective. For a formal definition of these errors, we refer the reader to \citet{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014, luNoteTypeErrors2018}. Take as an example the previous fictitious study on the relationship between anxiety and depression and imagine we decide to sample 50 individuals (sample size, n = 50) and to set the $\alpha$ level to 5\% and to perform a two-tailed test. On the basis of theoretical considerations, we expect the plausibly true correlation in the population to be quite strong and positive which we formalize as $\rho$ = .50. In order to evaluate the Type M and Type S errors in this research design, imagine repeating the same study many times with new samples drawn from the same population and, for each study, register the observed correlation (r) and the corresponding p-value.

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional a priori power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \citep{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitations techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \todo{(Anderson, 2019)}. The plausible effect size approximates the true effect, which is never known but can be thought as “that which would be observed in a hypothetical infinitely large sample” \citep[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description on plausible effect size, we refer the interested reader to \citet{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

Researchers can consider Type M and Type S errors in their practice by performing a \emph{design analysis} \citep{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}. Ideally, a design analysis should be performed when designing a study. In this phase it is specifically called \emph{prospective design analysis} and it can be used as a sample size planning strategy where statistical power is considered together with Type M and Type S errors. However, design analysis can also be beneficial to evaluate the inferential risks in studies that have already been conducted and where the study design is known. In these cases, Type M and Type S errors can support results interpretation by communicating the inferential risks in that research design. When design analysis happens at this later stage, it takes the name of \emph{retrospective design analysis}.

@@

We consider the study published in \emph{Science} by \citet{eisenbergerDoesRejectionHurt2003} entitled: “Does Rejection Hurt? An fMRI Study of Social Exclusion”. The research question originated from the observation that the Anterior Cingulate Cortex (ACC) is a region of the brain known to be involved in the experience of physical pain. Could pain from social stimuli, such as social exclusion, share similar neural underpinnings? To test this hypothesis, 13 participants were recruited and each one had to play a virtual game with other two players while undergoing functional Magnetic Resonance Imaging (fMRI). The other two players were fictitious, and participants were actually playing against a computer program. Players had to toss a virtual ball among each other in three conditions: social inclusion, explicit social exclusion and implicit social exclusion. In the social inclusion condition the participant regularly received the ball. In the explicit social exclusion condition the participant was told that, due to technical problems, he was not going to play that round. In the implicit social exclusion condition, the participant experienced being intentionally left out from the game by the other two players. At the end of the experiment, each participant completed a self-report measure regarding their perceived distress when they were intentionally left out by the other players. Considering only the implicit social exclusion condition, a correlation coefficient was estimated between the measure of distress and neural activity in the Anterior Cingulate Cortex. As suggested by the large and statistically significant correlation coefficient between perceived distress and activity in the ACC, \todo{r = .88, p < .005} \citep[p. 291]{eisenbergerDoesRejectionHurt2003}, authors concluded that social and physical pain seem to share similar neural underpinnings.

Before proceeding to the retrospective design analysis, we refer the interested reader to some background history regarding this study. This was one of the many studies included in the famous paper “Puzzlingly High Correlations in fMRI Studies of Emotion, Personality, and Social Cognition” \citep{vulPuzzlinglyHighCorrelations2009} which raised important issues regarding the analysis of neuroscientific data. In particular, this paper noted that the magnitude of correlation coefficients between fMRI measures and behavioral measures were beyond what could be considered plausible. We refer the interested reader also to the commentary by \citet{yarkoniBigCorrelationsLittle2009}, who noted that the implausibly high correlations in fMRI studies could be largely explained by the low statistical power of experiments.

A retrospective design analysis should start with thorough reasoning on the plausible size and direction of the effect under study. In order to produce valid inferences, a lot of attention should be devoted to this point by integrating external information. For the sake of this example, we turn to the considerations made by \citet{vulSuspiciouslyHighCorrelations2017} who \todo{considered} correlations between personality measures and neural activity to be likely around $\rho=.25$.  A correlation of $\rho=.50$ was deemed plausible but optimistic and a correlation of $\rho=.75$ was considered theoretically plausible but unrealistic.

@@

To perform a retrospective design analysis on the case study, we need information on the research design and the plausible effect size. Based on the previous considerations, we set the plausible effect size to be $\rho=.25$. Information on the sample size was not available in the original study \citep{eisenbergerDoesRejectionHurt2003} and was retrieved from \citet{vulPuzzlinglyHighCorrelations2009} to be $n=13$. The $\alpha$ level and the directionality of the test were not reported in the original study, so for the purpose of this example we will consider $\alpha=.05$  and a two-tailed test. Given this study design, what are the inferential risks in terms of effect size estimation?

We can use the R function \texttt{retro\_r()}, which inputs and outputs are displayed in Figure~\ref{fig:retro_r}. In this study, the statistical power is .13, that is to say, there is a 13\% probability to reject the null hypothesis, if an effect of at least $\rho=|.25|$ exists. Consider this point together with the results obtained in the experiment: r = .88, p < .005 \citep[p. 291]{eisenbergerDoesRejectionHurt2003}. It is clear that, even though the probability to reject the null hypothesis is low (power of 13\%), this event could happen. And when it does happen, it is tempting to believe that results are \todo{impressive} \citep{gelmanStatisticalCrisisScience2014}. However, this design comes with serious inferential risks for the estimation of effect sizes, which could be grasped by presenting Type M and Type S errors. A glance at their value communicates that it is not impossible to find a statistically significant result, but when it does happen, the effect sizes could be largely overestimated - Type M = 2.58 - and maybe even in the wrong direction - Type S = .03. The Type M error rate of \todo{around} 2.60 indicates that a statistically significant correlation is on average about two and a half times the plausible value. In other words, statistically significant results emerging in such a research design will on average overestimate the plausible correlation coefficient by 160\%. The Type S error of  .03 suggests that there is a three percent probability to find a statistically significant result in the opposite direction, in this example, a negative relationship.

@@

\begin{figure}[!h]
\centering
  \includegraphics[height=5cm]{screens/retro}
\caption{Input and Output of the function \texttt{retro\_r()} for retrospective design analysis. Case study: \citet{eisenbergerDoesRejectionHurt2003}. The plausible correlation coefficient is $\rho = .25$, the sample size is 13, and the statistical test is two-tailed.}\label{fig:retro_r}
\end{figure}

In this research design, the critical values above which a statistically significant result is declared correspond to $\rho=\pm .55$ (Figure~\ref{fig:retro_r}).  These values are highlighted in Figure~\ref{fig:Plot_winner} as the vertical lines in the sampling distribution of correlation coefficients under the null hypothesis. Notice that the plausible effect size lies in the region of acceptance of the null hypothesis. Therefore, it is impossible to simultaneously find a statistically significant result and estimate an effect close to the plausible one ($\rho=.25$). The figure represents the so called Winner’s curse: “the ‘lucky’ scientist who makes a discovery is cursed by finding an inflated estimate of that effect” \citep{buttonPowerFailureWhy2013}.

@@

For completeness, Figure~\ref{fig:Plot_scenarios} summarizes the relationship between statistical power, Type M and Type S errors as a function of sample size in three scenarios of plausible correlation coefficients. We display the three values that \citet{vulSuspiciouslyHighCorrelations2017} considered for correlations between fMRI measures and behavioral measures with different degrees of plausibility. An effect of $\rho=.75$ was deemed theoretically plausible but unrealistic, $\rho=.50$ was more plausible but optimistic, and $\rho=.25$ was more likely.  The curves illustrate a general point: Type M and Type S error increase with smaller sample sizes, smaller plausible effect sizes and lower statistical power. Also, the figure shows that statistical power, Type M and Type S errors are related to each other: as power increases, Type M and Type S errors decrease.

@@

A lot has been written regarding the automatic adoption of a conventional $\alpha$ level of 5\% \citep[e.g.,][]{gigerenzerNullRitualWhat2004, lakensJustifyYourAlpha2018}. This practice is increasingly discouraged, and researchers are invited to think about the best trade-off between $\alpha$ level and statistical power, considering the aim of the study and available resources. The $\alpha$ level impacts Type M and Type S errors as much as it impacts statistical power. Everything else equal, Type M error increases with decreasing $\alpha$ level (i.e., negative relationship), whereas Type S error decreases with decreasing $\alpha$ level (i.e., positive relationship). To further illustrate the relation between Type M error and $\alpha$ level, let us take as an example the previous case study with a sample of 13 participants, plausible effect size $\rho=.25$ and two-tailed test. Table~\ref{tab:Table_alpha} shows that by lowering the $\alpha$ level from 10\% to .10\%, the critical values move from \todo{$\rho=\pm .48$  to $\rho=\pm .80$}. This suggests that, with these new higher thresholds, the exaggeration of effects will be even more pronounced because effects have to be even larger to pass such higher critical values. Instead, the relationship between Type S error and $\alpha$ level can be clarified thinking that by lowering the statistical significance threshold, we are being more conservative to falsely reject the null hypothesis in general which implies that we are also being more conservative to falsely rejecting the null hypothesis in the wrong direction.

@@

\phantomsection
\addcontentsline{toc}{section}{Bibliography}
%\printbibliography
\bibliography{Paper_main}

@@

\appendix
\titleformat{\section}{\normalfont\Large\bfseries}{\setappendix}{0em}{}
\renewcommand\thefigure{\thesection.\arabic{figure}}
\renewcommand\theequation{\thesection.\arabic{equation}}
%----------------------------------------------------------------------------------%
%---------------------------       Appendix A      --------------------------------%
%----------------------------------------------------------------------------------%

@@

@@

\section{Session Information}

>>>>

%----    Basic packages    ----%
\documentclass{article}
\usepackage[T1] {fontenc} 		  % Font encoding
\usepackage [utf8] {inputenc}		% Encoding for the document
\usepackage[a4paper,includeheadfoot,left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}  % margin settings
\usepackage[british]{babel}

@@

%----    Other packages    ----%
\usepackage{tikz}
\usepackage{graphicx}   % for including graphics
\usepackage{amsmath}    % for math equation
\usepackage{bm}         % for bold math
\usepackage{csquotes}
\usepackage{enumitem}   % for enumarate and itemize
\setlist[enumerate]{font=\bfseries} % bold font for enumerate list
\setlist[enumerate]{itemsep=0ex}    % reduce space between items
\setlist[itemize]{itemsep=0ex}      % reduce space between items
\usepackage[labelfont=bf]{caption}  % caption
\usepackage{lineno}                 % rows number
\usepackage[doublespacing]{setspace}% line spacing
\usepackage{titlesec}               % redefine title format for appendix
\usepackage[style=apa, sortcites=true, sorting=nyt, backend=biber]{biblatex}  % bibliografia
\DeclareLanguageMapping{british}{british-apa}
\addbibresource{Paper_main.bib}
% \usepackage[natbibapa]{apacite}
% \bibliographystyle{apacite}

\usepackage{hyperref}    % ref between elements

%----    biber customs    ----%
% remove day-month
\AtEveryBibitem{%
  \clearfield{day}%
  \clearfield{month}%
  \clearfield{endday}%
  \clearfield{endmonth}%
}

% print url if no doi
% \DeclareSourcemap{
%   \maps[datatype=biber]{
%     \map[overwrite]{
%       \step[fieldsource=doi, final]
%       \step[fieldset=url, null]
%       \step[fieldset=eprint, null]
%     }
%   }
% }

@@

Psychological science is increasingly committed to scrutinize its published findings by promoting large-scale replication efforts \parencite{camererEvaluatingReplicabilityLaboratory2016, camererEvaluatingReplicabilitySocial2018, ebersoleManyLabsEvaluating2016, kleinInvestigatingVariationReplicability2014, kleinManyLabsInvestigating2018, opensciencecollaborationEstimatingReproducibilityPsychological2015}. In replication studies, the protocol of a previously published study is repeated as closely as possible with a new sample. Interestingly, many replication studies found smaller effects than originals, suggesting a sort of decline effect from original studies to replications \parencite{camererEvaluatingReplicabilitySocial2018, opensciencecollaborationEstimatingReproducibilityPsychological2015, schoolerTurningLensScience2014}. Among the many factors that could explain such decline, one relates to a feature of study design: statistical power \parencite{protzkoDeclineEffectsTypes2017}. In particular, it is plausible for original studies to have lower statistical power than their replications. When underpowered studies are analyzed using thresholds, such as statistical significance levels, effects passing such thresholds have to exaggerate the true effect size \parencite{buttonPowerFailureWhy2013, gelmanTypeErrorMight2017, ioannidisWhyMostDiscovered2008, ioannidisEmergenceLargeTreatment2013, laneEstimatingEffectSize1978}. Indeed, as will be extensively shown below, in underpowered studies only large effects correspond to values that can reject the null hypothesis and be statistically significant. As a consequence, if the original study was underpowerd and found an exaggerated estimate of the effect, the replication results will likely show a decline. The concept of statistical power finds its natural developement in the Neyman-Pearson framework of statistical inference and this is the framework that we adopt in this contribution. Further discussion on the Neyman and Perason approach and a comparison with Null Hypothesis Significance Testing (NHST) is available in \textcite{altoeEnhancingStatisticalInference2020} and \textcite{gigerenzerNullRitualWhat2004}.

Effect size exaggeration is a corollary consequence of low statistical power that might not be evident at first. Indeed, by definition, statistical power focuses on statistical significance, not effect size estimation (i.e., statistical power is the probability to find a statistically significant result if an effect of a certain size actually exists). However, effect size estimation and statistical significance are closely related. This point can be highlighted estimating the Type M (magnitude) and Type S (sign) errors characterizing a study design \parencite{gelmanPowerCalculationsAssessing2014}. Given a study design (i.e., sample size, statistical test directionality, and $\alpha$ level), Type M error indicates the factor by which a statistically significant effect would be, on average, exaggerated. Type S error indicates the probability to find a statistically significant effect in the opposite direction to the one considered plausible. It is evident that both errors are defined starting from a reasoned guess on the plausible magnitude and direction of the effect under study, which is called \emph{plausible effect size} \parencite{gelmanPowerCalculationsAssessing2014}. The analysis that researchers perform to consider Type M and Type S errors in their research practice is called \emph{design analysis}, given the special focus posed into considering the design of a study \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

Why do these errors matter? The exaggeration of effect sizes, in the right or in the wrong direction, has important consequences on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, which could turn out to be underpowered if they do not take into account possible inflations of the original effect \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanEmbracingVariationAccepting2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

Type M (magnitude) and Type S (sign) errors are not widely known in the psychological research community but their consideration during the research process has the potential to improve current research practices, for example, by increasing the awarness that design choices have on possible studies' results. In a previous work we illustrated Type M and Type S errors using Cohens’\emph{d} as a measure of effect size \parencite{altoeEnhancingStatisticalInference2020}. In this contribution we aim to to further increase familiarity with Type M and Type S errors, considering another common effect size measures in psychology: Pearson correlation coefficient, $\rho$.  In the following paragraphs we discuss what Type M and Type S errors are and how to perform a design analysis using the ad-hoc R functions that we developed.

@@

Pearson correlation coefficient is a standardized effect size measure indicating the strength and the direction of the relationship between two continuous variables \parencite{cohenStatisticalPowerAnalysis1988, ellisEssentialGuideEffect2010}. Even though the correlation coefficient is widely known, we briefly go over its main features using an example. Imagine that we were interested to measure the relationship between anxiety and depression in a population and we plan a study with \emph{n} participants, where, for each participant, we measure the level of anxiety (i.e., variable X) and the level of depression (i.e., variable Y). At the end of the study we will have n pairs of values X and Y. The correlation coefficient helps us answer the questions: how strong is the relationship between anxiety and depression in this population? Is the relationship positive or negative? Correlation ranges from -1 to +1, indicating respectively two extreme scenarios of perfect negative relationship and perfect positive relationship. Since the correlation coefficient is a dimensionless number, it is a signal to noise ratio where the signal is given by the covariance between the two variables ($cov(x,y)$) and the noise is expressed by the product between the standard deviations of the two variables ($S_x  S_y$; see Formula\ref{eq:correlation}). Following the conventional standards, in this contribution we will use the symbol $\rho$ to indicate the correlation in the population and the symbol \emph{r} to indicate the value measured in a sample.
\begin{equation}
\label{eq:correlation}
r=\frac{cov(x,y)}{S_x S_y}.
\end{equation}

Magnitude and sign are two important features characterizing Pearson correlation coefficient and effect size measures in general. And, when estimating effect sizes, errors could be committed exactly regarding these two aspects. \textcite{gelmanPowerCalculationsAssessing2014} introduced two indexes to quantify these risks:
\begin{itemize}
  \item{Type M error, where M stands for magnitude, is also called Exaggeration Ratio - the factor by which a statistically significant effect is on average exaggerated.}
  \item{Type S error, sign - the probability to find a statistically significant result in the opposite direction to the plausible one.}
\end{itemize}

How are these errors computed? In the next paragraphs we approach this question preferring an intuitive perspective. For a formal definition of these errors, we refer the reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014, luNoteTypeErrors2018}. Take as an example the previous fictitious study on the relationship between anxiety and depression and imagine we decide to sample 50 individuals (sample size, n = 50) and to set the $\alpha$ level to 5\% and to perform a two-tailed test. On the basis of theoretical considerations, we expect the plausibly true correlation in the population to be quite strong and positive which we formalize as $\rho$ = .50. In order to evaluate the Type M and Type S errors in this research design, imagine repeating the same study many times with new samples drawn from the same population and, for each study, register the observed correlation (r) and the corresponding p-value.

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional a priori power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitations techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \todo{(Anderson, 2019)}. The plausible effect size approximates the true effect, which is never known but can be thought as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description on plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

Researchers can consider Type M and Type S errors in their practice by performing a \emph{design analysis} \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}. Ideally, a design analysis should be performed when designing a study. In this phase it is specifically called \emph{prospective design analysis} and it can be used as a sample size planning strategy where statistical power is considered together with Type M and Type S errors. However, design analysis can also be beneficial to evaluate the inferential risks in studies that have already been conducted and where the study design is known. In these cases, Type M and Type S errors can support results interpretation by communicating the inferential risks in that research design. When design analysis happens at this later stage, it takes the name of \emph{retrospective design analysis}.

@@

We consider the study published in \emph{Science} by \textcite{eisenbergerDoesRejectionHurt2003} entitled: “Does Rejection Hurt? An fMRI Study of Social Exclusion”. The research question originated from the observation that the Anterior Cingulate Cortex (ACC) is a region of the brain known to be involved in the experience of physical pain. Could pain from social stimuli, such as social exclusion, share similar neural underpinnings? To test this hypothesis, 13 participants were recruited and each one had to play a virtual game with other two players while undergoing functional Magnetic Resonance Imaging (fMRI). The other two players were fictitious, and participants were actually playing against a computer program. Players had to toss a virtual ball among each other in three conditions: social inclusion, explicit social exclusion and implicit social exclusion. In the social inclusion condition the participant regularly received the ball. In the explicit social exclusion condition the participant was told that, due to technical problems, he was not going to play that round. In the implicit social exclusion condition, the participant experienced being intentionally left out from the game by the other two players. At the end of the experiment, each participant completed a self-report measure regarding their perceived distress when they were intentionally left out by the other players. Considering only the implicit social exclusion condition, a correlation coefficient was estimated between the measure of distress and neural activity in the Anterior Cingulate Cortex. As suggested by the large and statistically significant correlation coefficient between perceived distress and activity in the ACC, \todo{r = .88, p < .005} \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}, authors concluded that social and physical pain seem to share similar neural underpinnings.

Before proceeding to the retrospective design analysis, we refer the interested reader to some background history regarding this study. This was one of the many studies included in the famous paper “Puzzlingly High Correlations in fMRI Studies of Emotion, Personality, and Social Cognition” \parencite{vulPuzzlinglyHighCorrelations2009} which raised important issues regarding the analysis of neuroscientific data. In particular, this paper noted that the magnitude of correlation coefficients between fMRI measures and behavioral measures were beyond what could be considered plausible. We refer the interested reader also to the commentary by \textcite{yarkoniBigCorrelationsLittle2009}, who noted that the implausibly high correlations in fMRI studies could be largely explained by the low statistical power of experiments.

A retrospective design analysis should start with thorough reasoning on the plausible size and direction of the effect under study. In order to produce valid inferences, a lot of attention should be devoted to this point by integrating external information. For the sake of this example, we turn to the considerations made by \textcite{vulSuspiciouslyHighCorrelations2017} who \todo{considered} correlations between personality measures and neural activity to be likely around $\rho=.25$.  A correlation of $\rho=.50$ was deemed plausible but optimistic and a correlation of $\rho=.75$ was considered theoretically plausible but unrealistic.

@@

To perform a retrospective design analysis on the case study, we need information on the research design and the plausible effect size. Based on the previous considerations, we set the plausible effect size to be $\rho=.25$. Information on the sample size was not available in the original study \parencite{eisenbergerDoesRejectionHurt2003} and was retrieved from \textcite{vulPuzzlinglyHighCorrelations2009} to be $n=13$. The $\alpha$ level and the directionality of the test were not reported in the original study, so for the purpose of this example we will consider $\alpha=.05$  and a two-tailed test. Given this study design, what are the inferential risks in terms of effect size estimation?

We can use the R function \texttt{retro\_r()}, which inputs and outputs are displayed in Figure~\ref{fig:retro_r}. In this study, the statistical power is .13, that is to say, there is a 13\% probability to reject the null hypothesis, if an effect of at least $\rho=|.25|$ exists. Consider this point together with the results obtained in the experiment: r = .88, p < .005 \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}. It is clear that, even though the probability to reject the null hypothesis is low (power of 13\%), this event could happen. And when it does happen, it is tempting to believe that results are \todo{impressive} \parencite{gelmanStatisticalCrisisScience2014}. However, this design comes with serious inferential risks for the estimation of effect sizes, which could be grasped by presenting Type M and Type S errors. A glance at their value communicates that it is not impossible to find a statistically significant result, but when it does happen, the effect sizes could be largely overestimated - Type M = 2.58 - and maybe even in the wrong direction - Type S = .03. The Type M error rate of \todo{around} 2.60 indicates that a statistically significant correlation is on average about two and a half times the plausible value. In other words, statistically significant results emerging in such a research design will on average overestimate the plausible correlation coefficient by 160\%. The Type S error of  .03 suggests that there is a three percent probability to find a statistically significant result in the opposite direction, in this example, a negative relationship.

@@

\begin{figure}[!h]
\centering
  \includegraphics[height=5cm]{screens/retro}
\caption{Input and Output of the function \texttt{retro\_r()} for retrospective design analysis. Case study: \textcite{eisenbergerDoesRejectionHurt2003}. The plausible correlation coefficient is $\rho = .25$, the sample size is 13, and the statistical test is two-tailed.}\label{fig:retro_r}
\end{figure}

In this research design, the critical values above which a statistically significant result is declared correspond to $\rho=\pm .55$ (Figure~\ref{fig:retro_r}).  These values are highlighted in Figure~\ref{fig:Plot_winner} as the vertical lines in the sampling distribution of correlation coefficients under the null hypothesis. Notice that the plausible effect size lies in the region of acceptance of the null hypothesis. Therefore, it is impossible to simultaneously find a statistically significant result and estimate an effect close to the plausible one ($\rho=.25$). The figure represents the so called Winner’s curse: “the ‘lucky’ scientist who makes a discovery is cursed by finding an inflated estimate of that effect” \parencite{buttonPowerFailureWhy2013}.

@@

For completeness, Figure~\ref{fig:Plot_scenarios} summarizes the relationship between statistical power, Type M and Type S errors as a function of sample size in three scenarios of plausible correlation coefficients. We display the three values that \textcite{vulSuspiciouslyHighCorrelations2017} considered for correlations between fMRI measures and behavioral measures with different degrees of plausibility. An effect of $\rho=.75$ was deemed theoretically plausible but unrealistic, $\rho=.50$ was more plausible but optimistic, and $\rho=.25$ was more likely.  The curves illustrate a general point: Type M and Type S error increase with smaller sample sizes, smaller plausible effect sizes and lower statistical power. Also, the figure shows that statistical power, Type M and Type S errors are related to each other: as power increases, Type M and Type S errors decrease.

@@

A lot has been written regarding the automatic adoption of a conventional $\alpha$ level of 5\% \parencite[e.g.,][]{gigerenzerNullRitualWhat2004, lakensJustifyYourAlpha2018}. This practice is increasingly discouraged, and researchers are invited to think about the best trade-off between $\alpha$ level and statistical power, considering the aim of the study and available resources. The $\alpha$ level impacts Type M and Type S errors as much as it impacts statistical power. Everything else equal, Type M error increases with decreasing $\alpha$ level (i.e., negative relationship), whereas Type S error decreases with decreasing $\alpha$ level (i.e., positive relationship). To further illustrate the relation between Type M error and $\alpha$ level, let us take as an example the previous case study with a sample of 13 participants, plausible effect size $\rho=.25$ and two-tailed test. Table~\ref{tab:Table_alpha} shows that by lowering the $\alpha$ level from 10\% to .10\%, the critical values move from \todo{$\rho=\pm .48$  to $\rho=\pm .80$}. This suggests that, with these new higher thresholds, the exaggeration of effects will be even more pronounced because effects have to be even larger to pass such higher critical values. Instead, the relationship between Type S error and $\alpha$ level can be clarified thinking that by lowering the statistical significance threshold, we are being more conservative to falsely reject the null hypothesis in general which implies that we are also being more conservative to falsely rejecting the null hypothesis in the wrong direction.

@@

\phantomsection
\addcontentsline{toc}{section}{Bibliography}
\printbibliography
%\bibliography{Paper_main}

@@

\appendix
\titleformat{\section}{\normalfont\Large\bfseries}{\setappendix}{0em}{}
\renewcommand\thefigure{\thesection.\arabic{figure}}
\renewcommand\theequation{\thesection.\arabic{equation}}
%----------------------------------------------------------------------------------%
%---------------------------       Appendix A      --------------------------------%
%----------------------------------------------------------------------------------%
\begin{refsection}

@@

\printbibliography[heading=subbibliography, title = Appendix References ]
\end{refsection}

@@

\section*{Session Information}
